---
title: "Practical-Homework-2"
author: "Hamda Hassan"
date: "Due April 26, 2024"
output: html_document
---


```{r} 
library(tidyverse) 
library(e1071) 
library(ISLR2) 
library(caret) 
library(ggplot2)
```


## LOAD DATA

Summary: The dataset was loaded directly from the public GitHub repository. It contains health, 
demographic, and behavioral variables used to predict diabetes.
```{r}
url <- "https://raw.githubusercontent.com/mendible/5322/main/Homework%202/nhis_2022.csv"
data <- read.csv(url)
```


## INITIAL DATA CHECK

Summary: Initial exploration was performed to understand the structure, variables, and distribution of the dataset. Additional checks were used to identify missing values and detect unrealistic or placeholder values (e.g., extreme values like 999), which helped guide the data cleaning process.
```{r}
str(data)
colnames(data)
head(data)
summary(data)
table(data$DIABETICEV)
```


```{r}
# Count missing values in each column
colSums(is.na(data))

# Check suspicious high values
summary(data$AGE)
summary(data$BMICALC)
summary(data$MOD10DMIN)
summary(data$SODAPNO)

```

Summary: Histograms were used to visualize the distribution of key variables. This helped identify skewness, outliers, and unrealistic values that needed to be addressed during cleaning.
```{r}
hist(data$AGE)
hist(data$BMICALC)
hist(data$MOD10DMIN)
hist(data$SODAPNO)
```


Summary: Simple checks were used to compare average values of key variables across diabetes groups. This gave an early sense of potential relationships in the data.
```{r}
aggregate(AGE ~ DIABETICEV, data = data, mean, na.rm = TRUE)
aggregate(BMICALC ~ DIABETICEV, data = data, mean, na.rm = TRUE)
```


## TARGET VARIABLE

Summary: The target variable was created by recoding DIABETICEV into a binary outcome (1 = diabetes, 0 = no diabetes). Unclear responses were treated as missing, and the variable was converted to a factor for classification.
```{r}
data <- data %>%
  mutate(
    diabetes = case_when(
      DIABETICEV == 2 ~ 1,  # YES = 1
      DIABETICEV == 1 ~ 0,  # NO = 0
      TRUE ~ NA_real_
    ),
    diabetes = as.factor(diabetes)
  )
```


## CLEANING + Build Model

Summmary: The data was cleaned by removing invalid placeholder values and restricting the sample to adults (18+). A subset of relevant predictors was selected, extreme soda values were capped, categorical variables were formatted, and missing observations were removed to create the final modeling dataset.
```{r}
model_data <- data %>%
  filter(AGE >= 18) %>%
  select(diabetes, AGE, SEX, BMICALC, MOD10DMIN, SODAPNO) %>%
  mutate(across(
    -diabetes,
    ~ifelse(. %in% c(7,8,9,96,97,98,99,996,997,998,999), NA, .)
  )) %>%
  mutate(
    SODAPNO = ifelse(SODAPNO > 30, 30, SODAPNO),
    SEX = factor(SEX, levels = c(1, 2),
                 labels = c("Male", "Female"))
  ) %>%
  drop_na()
```


## CHECKING FOR MISTAKES

Summary: These checks were used to confirm that the cleaned dataset looked reasonable before modeling. I checked the final variable summaries, diabetes class balance, and sex distribution to make sure there were no obvious cleaning issues left.
```{r}
summary(model_data)
table(model_data$diabetes)
table(model_data$SEX)

```

## TRAIN / TEST SPLIT FIRST

Summary: The cleaned dataset was split into training and testing sets, with 80% used for training and 20% held out for evaluation. A random seed was set so the results could be reproduced.
```{r}
set.seed(123)

train_index <- sample(1:nrow(model_data), size = 0.8 * nrow(model_data))

train_raw <- model_data[train_index, ]
test_raw  <- model_data[-train_index, ]
```



## DUMMY VARIABLES + SCALE AFTER SPLIT

Summary: Categorical variables were converted into dummy variables, and all predictors were scaled to ensure the SVM models performed correctly. These steps were applied after the train/test split to prevent data leakage.
```{r}
# Convert predictors only into dummy variables
dummies <- dummyVars(~ ., data = train_raw[, -1])

train_matrix <- predict(dummies, newdata = train_raw[, -1])
test_matrix  <- predict(dummies, newdata = test_raw[, -1])

# Scale numeric predictors after dummy encoding
preproc <- preProcess(train_matrix, method = c("center", "scale"))

train_scaled <- predict(preproc, train_matrix)
test_scaled  <- predict(preproc, test_matrix)

train <- data.frame(diabetes = train_raw$diabetes, train_scaled)
test  <- data.frame(diabetes = test_raw$diabetes, test_scaled)

# Reattach response variable and ensure it is a factor
train$diabetes <- factor(train$diabetes, levels = c("0", "1"))
test$diabetes  <- factor(test$diabetes, levels = c("0", "1"))
```

## Tests

Summary: Final checks were performed to confirm that the training and testing datasets were properly structured, had consistent factor levels, and contained no missing values.

```{r}
str(train)
str(test)

```

```{r}
levels(train$diabetes)
levels(test$diabetes)
```

```{r}
anyNA(train)
anyNA(test)

```


## CLASS WEIGHTS (IMPORTANT)

Summary: Class weights were applied to address the imbalance in the dataset, where far fewer individuals have diabetes. Higher weight was given to the diabetes class to improve the model's ability to correctly detect positive cases.
```{r}
class_counts <- table(train$diabetes)

class_weights <- c(
  "0" = 1,
  "1" = as.numeric(class_counts["0"] / class_counts["1"])
)

print(class_counts)
print(class_weights)
```


## BASE MODELS

Summary: Three SVM models (radial, polynomial, and linear) were trained using the weighted dataset. Each model was evaluated on the test set using accuracy, sensitivity, and specificity to compare performance across different kernel types.


Linear
```{r}
svm_linear <- svm(
  diabetes ~ .,
  data = train,
  kernel = "linear",
  cost = 1,
  class.weights = class_weights
)

pred_linear_base <- predict(svm_linear, test)
cm_linear_base <- confusionMatrix(pred_linear_base, test$diabetes, positive = "1")
cm_linear_base
```


Radial
```{r}
svm_radial <- svm(
  diabetes ~ .,
  data = train,
  kernel = "radial",
  cost = 1,
  gamma = 0.05,
  class.weights = class_weights
)

pred_radial_base <- predict(svm_radial, test)
cm_radial_base <- confusionMatrix(pred_radial_base, test$diabetes, positive = "1")
cm_radial_base
```

Polynomial
```{r}
set.seed(123)
train_small <- train[sample(1:nrow(train), min(5000, nrow(train))), ]

svm_poly <- svm(
  diabetes ~ .,
  data = train_small,
  kernel = "polynomial",
  degree = 2,
  cost = 1,
  class.weights = class_weights
)

pred_poly_base <- predict(svm_poly, test)
cm_poly_base <- confusionMatrix(pred_poly_base, test$diabetes, positive = "1")
cm_poly_base
```




## TUNING

Summary: The radial SVM was tuned using cross-validation to find the best combination of cost and gamma values. Multiple parameter values were tested, and the model with the best performance on validation data was selected and evaluated on the test set.
```{r}
tuned_radial <- tune(
  svm,
  diabetes ~ .,
  data = train,
  kernel = "radial",
  ranges = list(
    cost = c(0.5, 1, 2, 5),
    gamma = c(0.01, 0.05, 0.1)
  ),
  tunecontrol = tune.control(cross = 3),
  class.weights = class_weights
)

best_radial <- tuned_radial$best.model

pred_radial <- predict(best_radial, test)
cm_radial_tuned <- confusionMatrix(pred_radial, test$diabetes, positive = "1")
cm_radial_tuned

```

## MODEL COMPARISON

Summary: Model performance was summarized in a comparison table using accuracy, sensitivity, and specificity. This allowed for a clear comparison across all SVM kernels, including the tuned radial model.
```{r}
results <- data.frame(
  Model = c("Linear", "Radial", "Polynomial", "Radial (Tuned)"),
  Accuracy = c(
    cm_linear_base$overall["Accuracy"],
    cm_radial_base$overall["Accuracy"],
    cm_poly_base$overall["Accuracy"],
    cm_radial_tuned$overall["Accuracy"]
  ),
  Sensitivity = c(
    cm_linear_base$byClass["Sensitivity"],
    cm_radial_base$byClass["Sensitivity"],
    cm_poly_base$byClass["Sensitivity"],
    cm_radial_tuned$byClass["Sensitivity"]
  ),
  Specificity = c(
    cm_linear_base$byClass["Specificity"],
    cm_radial_base$byClass["Specificity"],
    cm_poly_base$byClass["Specificity"],
    cm_radial_tuned$byClass["Specificity"]
  )
)

results

```


## SVM DECISION BOUNDARY PLOT

Summary: A two-variable SVM decision boundary was visualized using BMI and physical activity to illustrate how the model separates classes. The plot shows that these variables alone do not clearly separate diabetes cases, highlighting the need for more complex models.
```{r}

set.seed(123)

plot_data <- model_data %>%
  select(diabetes, BMICALC, MOD10DMIN) %>%
  filter(MOD10DMIN <= 200) %>%   # key fix
  sample_n(2000)

# Train SVM
svm_2d <- svm(
  diabetes ~ BMICALC + MOD10DMIN,
  data = plot_data,
  kernel = "radial",
  cost = 1,
  gamma = 0.05
)

# Plot
plot(
  svm_2d, plot_data,
  main = "SVM Decision Boundary (BMI vs Physical Activity)",
  xlab = "BMI",
  ylab = "Moderate Activity (Minutes)",
  col = c("lightblue", "pink")
)


```


## BMI vs Diabetes

Summary: A boxplot of BMI by diabetes status shows a clear upward shift for individuals with diabetes, indicating that BMI is a strong predictor.
```{r}
ggplot(model_data, aes(x = diabetes, y = BMICALC, fill = diabetes)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "BMI Distribution by Diabetes Status",
    x = "Diabetes (0 = No, 1 = Yes)",
    y = "BMI"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```


## Physical Activity vs Diabetes

Summary: A boxplot of physical activity shows significant overlap between groups, suggesting it is a weaker predictor when considered on its own.
```{r}
ggplot(model_data, aes(x = diabetes, y = MOD10DMIN, fill = diabetes)) +
  geom_boxplot(alpha = 0.7) +
  coord_cartesian(ylim = c(0, 200)) + 
  labs(
    title = "Physical Activity by Diabetes Status",
    x = "Diabetes (0 = No, 1 = Yes)",
    y = "Moderate Activity (Minutes per Day)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```































































